Supplement to ``{\mytitle}''

### 2022-12-17 si/si.Rnw
Changed table parameters, worked on reference and supp

%%% JASA. Attention Mac Users. The latest version of the pdfTex compiler for Mac (pdfTeX-1.40.11) outputs files not currently supported by ScholarOne Manuscripts. To resolve this and have a PDF file that will convert properly, the PDF file will need to be distilled to Adobe version 1.4 or earlier. To create a PDF file that uploads correctly, place the command \pdfminorversion=4 in the preamble (before the begin{document}) of your LaTeX file and run pdfLaTeX again.

@@

%Draft: \today

@@

@@

We use the reported cases at the start of the pandemic to approximate the number of Asymptomatic, Infectious, and Recovered individuals in each department $u \in 1:U$ using the same approximation as provided in Eq.~34.
% TODO: make sure it aligns with \myeqref{eq:model3:approx}.
The susceptible compartment is initialized so that the sum $S_{u}(0) + I_u(0) + A_u(0) + \sum_k R_{u, k}(0) = \text{population}_u$.
The bacteria compartment is then initialized using Eq.~\myeqref{eq:model3:b0}:
\begin{equation}
  B_u(0) = \big[1 + \seasAmplitude \big(\Binit_u)^r \big] D_i \, \Wshed \big[ I_{u}(0)+ \asymptomRelativeShed A_{u}(0) \big]\label{eq:model3:b0},
\end{equation}
where $\Binit_u \in (0, 1)$ is an initial value parameter that allows some flexibility in determining the initial state of the bacteria compartment.

This measurement model differs from that used by \citet{lee20}, who fit the model in two stages: epidemic and endemic phases.
Although their text and supplement material do not explicitly describe the measurement model used, inspection of the code provided with their submission suggests a change in measurement model between the epidemic and endemic phases.
In the file \code{choleraEqsPublic.py}, \citet{lee20} create several functions, where each function represents a set of coupled differential equations that could potentially be used to model cholera incidence data.
Each function returns a vector (or \code{numpy} array) that represents the change in each state variable for a single time step, including the variable \code{dC}, which tracks the number of new infections and is used to obtain the reported case counts.
Following their comments and code, it appears that the function \code{choleraEqs10WithoutVaccinationNetwork} was used to describe the dynamics of the epidemic phase, and \code{choleraEqs11WithoutVaccinationNetwork} was used for the endemic stage (see Sec~\ref{sec:mod2rep} for more details).
Because their models were fit using least squares, the code in these functions suggest that the measurement model for the epidemic phase is
\begin{equation}
  \label{model2:meas}
  Y_{u, n} \mid \Delta N_{EI}(n) = z(n) \sim \text{N}\left(\reportRate z(n), \obsOverdispersion^2 \right),
\end{equation}
which is similar to our measurement model, the primary difference being that the measurement model is applied to raw case counts rather than log-transformed case counts.
The measurement model for the endemic phase modifies the epidemic model by counting both asymptomatically infected (A) and symptomatically infected (I) individuals in the case counts:
\begin{equation}
  \label{model2:meas}
  Y_{u, n} \mid \Delta N_{EI}(n) = z_1(n), \Delta N_{EA}(n) = z_2(n) \sim \text{N}\left(\reportRate (z_1(n) + z_2), \obsOverdispersion^2 \right),
\end{equation}
where the notation for $\Delta N_{EA}(n)$ is similar to $\Delta N_{EI}(n)$, described above.

This measurement model is a minor change from that used by \cite{lee20}, which allowed for a change in the reporting rate on January 1st, 2018.
The fitted values of the reporting rate---before and after January 2018---were $0.97$ and $0.097$, respectively.
This major change in reporting rate alone could explain why Model~3 of \cite{lee20} failed to predict the eradication of cholera.
An overnight change from near perfect to almost non-existent reporting forces the model to explain the observed reduction in reported cases as a decrease in the reporting of cases, rather than a decrease in the prevalence of cholera.
This shift was justified by a "change of the case definition that occurred on January 1st, 2018";
this claim was not cited, and we could find no evidence that such a drastic change in the reporting rate would be warranted.
We therefore do not allow a change in reporting rate when fitting Model~3.

@@

@@

There are minor differences between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20sup}.
% such as a delayed second wave in the epidemic period and a larger initial wave during the endemic period.
While the discrepancy appears minor, the deterministic nature of Model~2 implies that an exact replication of model trajectories should be possible.
In this case, these discrepancies may be attributed to model implementation and small differences in model parameters.
Details about the measurement models and how latent states were initialized for the epidemic model were not provided by \citet{lee20sup} and therefore these details must be inferred by looking at the provided source code.
According to repository comments, the files \code{fit\allowbreak In\allowbreak Pieces\allowbreak 3params\allowbreak Clean\allowbreak May2019\allowbreak Public.py} and \code{fit\allowbreak In\allowbreak Pieces\allowbreak Mu\allowbreak With\allowbreak Frac\allowbreak Sus\allowbreak Fixed\allowbreak All\allowbreak Infections\allowbreak Public.py} were used to fit the epidemic and endemic phases of the model respectively, although it is apparent that these exact files were not used to obtain the reported results.
These files contain errors that making them impossible to run, and therefore these files cannot be the code that generated the reported model fits\footnote{One example of why the code cannot be run that the file loads functions from a non-extant file named \code{choleraEqs.py} in line 13 rather than \code{cholera\allowbreak Eqs\allowbreak Public.py}.}.
The inability to replicate the results by \citet{lee20} by running the provided source code makes checking whether or not a our numeric implementation faithfully represents their results very difficult.
% The difficulty of reproducing the result of \citet{lee20} is coupled with a misunderstanding of how these results were originally obtained.

\jwc{PROBABLY DELETE THIS NEXT BIT, BUT IT'S USEFUL TO DOCUMENT IT FOR NOW...}
One concern that arises when inspecting the files \code{fit\allowbreak In\allowbreak Pieces3\allowbreak params\allowbreak Clean\allowbreak May\allowbreak 2019\allowbreak Public.py} and \code{fit\allowbreak In\allowbreak Pieces\allowbreak Mu\allowbreak With\allowbreak Frac\allowbreak Sus\allowbreak Fixed\allowbreak All\allowbreak Infections\allowbreak Public.py} is that the output of these files (which contains the estimated parameters) was not referenced/used in either the file that generates the sensitivity analysis or the file that generates the model simulations; instead, alternative (non-extant) files with similar names are loaded, suggesting parameters that were used to generate the final results were not a direct result of the estimated parameters.
Furthermore, inspection of the script \code{simulate\allowbreak Haiti\allowbreak Full\allowbreak Time\allowbreak In\allowbreak Pieces\allowbreak All\allowbreak Vaccination\allowbreak Scenarios\allowbreak Single\allowbreak Parameter\allowbreak Vector\allowbreak All\allowbreak Infections\allowbreak Public.py}---which is used to obtain simulations for each of the vaccination scenarios---results in a number of concerns.
First, the script uses parameters from the pickle file \code{least\allowbreak Squares\allowbreak Data\allowbreak UpTo\allowbreak 2019\allowbreak /fit\allowbreak Epidemic\allowbreak Piece\_results\allowbreak Matrix\_18\allowbreak Mar\allowbreak 2019.pickle}, which only appears in the repository as a commented out line of code in the file \code{generate\allowbreak Sets\allowbreak Of\allowbreak Parameters\allowbreak For\allowbreak Sensitivity\allowbreak Analysis\allowbreak Public.py}, suggesting that the set of parameters used to simulate each vaccination scenario was not the most recent set of parameters generated by the sensitivity analysis.
Second, the script appears to use the set of differential equations from the function \code{cholera\allowbreak Eqs11\allowbreak With\allowbreak Vaccination\allowbreak Network} for the epidemic phase, even though parameter values for this phase were estimated using \code{cholera\allowbreak Eqs10\allowbreak With\allowbreak Vaccination\allowbreak Network}.
It is possible, however, that we simply are interpreting the code incorrectly and that these concerns are unwarranted.
Furthermore, because the provided code is non-functional, it is possible that the code that was actually used to obtain the reported results does not include these potential errors.
This highlights the importance of providing the exact source code that was used to obtain the reported results of any analysis.

Another potential explanation for the discrepancy between our simulations from Model~2 and those of \citet{lee20} is that the parameters that we used are only approximately the same as those used by \citet{lee20sup}. For example, the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely had values that were far more precise.
Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.
The collective effect of these small differences in model parameters likely will result in differences in model trajectories.
This issue could have been mitigated if the saved files containing estimated model parameters were included as part of the code submission.

@@

lee3 <- lee3_spatPomp()
lee3_sims <- bake("../model3/si_lee3Sims.rds", {
  simulate(lee3, nsim = NSIM, seed = 321, format = 'data.frame')
})

@@

\section{Calibrating Model~3 to observed cases}

# TODO: replace with bake
# haiti3_orig_fit <- readRDS(save_file)

In an attempt to determine why the model is underperforming in some departments, we plot the conditional log likelihoods of each observation given the previous observations for each department in Fig.~\ref{fig:condLL}.
This figure reveals that the fitted model has difficult time explaining certain features of the data.
For example, many departments (in particular Sud) have observations with lower conditional log-likelihoods near October 2016 than at other time points.
Further investigation of the model output reveals that the model is struggling to explain the sudden surge in cholera cases that occurred at this time, which coincides with the time that Hurricane Matthew struck Haiti.
While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \jwc{TODO: CITE}.
This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 that would increase the transmission between environmental cholera and humans for Sud and Grand'Anse.
The effect of the hurricane on cholera transmission is assumed to have an exponential decay, where the magnitude is determined by $\Whur{u}$ and the duration of the effect determined by $\hHur{u}$.

t(bpf_out@block.cond.loglik) %>%
  as.data.frame() %>%
  mutate(t = bpf_out@times) %>%
  pivot_longer(
    cols = -t,
    values_to = 'likelihood',
    names_prefix = "V",
    names_to = 'Unit'
  ) %>%
  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
  ggplot(aes(x = date, y = likelihood)) +
  geom_line() +
  geom_point() +
  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
             nrow = 2) +
  theme_bw() +
  theme(axis.title.x = element_blank()) +
  ylab("Conditional log-likelihoods")
@
\caption{\label{fig:condLL}Conditional log-likelihoods of Model~3 prior to the inclusion of parameters for initial values and Hurricane Matthew.}
\end{figure}

Other points of concern include the observations at or near the start of the recorded cases, for example, in Nord-Ouest and Sud-Est.
These low conditional log-likelihoods are a result of having zero or near zero confirmed cases in the first week of observed data, followed by a rapid growth of cases.
The model cannot explain the rapid growth of cholera in the absence of starting cases because the initialization scheme forces departments with no recorded cases in the first week to have hidden states $I_{u0}(0) = A_{u0}(0) = B_{u}(0) = 0$.
In reality, however, it is possible to have some infected or asymptomatic individuals in a department when no cases are recorded due to under-reporting.
This may be particularly true at the onset of an infectious disease outbreak, as lower public awareness and inadequate testing infrastructure may result in many cases going undetected \jwc{TODO: CITE}.
We therefore include and estimate $I_{0u}$ that determines the initial number of Infected individuals at the start of the pandemic for the departments Grande'Anse, Nord-Ouest, and Sud-Est;
each of these departments have both low starting case counts followed by rapid growth, and low initial values of the conditinoal log-likelihood.

\begin{figure}[ht]
<<finalCondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
t(bpf_out@block.cond.loglik) %>%
  as.data.frame() %>%
  mutate(t = bpf_out@times) %>%
  pivot_longer(
    cols = -t,
    values_to = 'likelihood',
    names_prefix = "V",
    names_to = 'Unit'
  ) %>%
  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
  ggplot(aes(x = date, y = likelihood)) +
  geom_line() +
  geom_point() +
  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
             nrow = 2) +
  theme_bw() +
  theme(axis.title.x = element_blank()) +
  ylab("Conditional log-likelihoods")
@
\caption{\label{fig:finalCondLL}Conditional log-likelihoods of Model~3 after adding and estimating the parameters for initial values and Hurricane Matthew.}
\end{figure}

Recall that the filtering distribution for the calibrated version of Model~3 at time $t_k$ is defined as the distribution of the latent state at time $t_k$ given the data from times $t_{1}:t_{k}$, i.e. $f^{(3)}_{\bm{X}_k|\bm{Y}_{1:k}}(\bm{x}_{k} | \bm{y}^*_{1:k} ; \hat\theta)$.
In general, one may expect simulations from the filtering distribution of a model with a good statistical fit to result in hidden states that are highly consistent with the observed data because the filtering distribution is conditioned on the observed data.
Fig.~\ref{fig:h3Sus} shows the percentage of individuals that are in the susceptible compartment from various simulations of the model:
simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
This figure shows that simulations from initial conditions tends to result in a much more rapid depletion of susceptible individuals at the start of the epidemic than simulations from the filtering distribution, suggesting the calibrated model has a propensity to predict larger outbreaks than what is typically seen in the data.
This also appears true in later stages of the time period of observed data, where the data suggests that there is a replenishment of the susceptible compartment, but the Model simulations typically retain a smaller proportion of susceptible individuals.
This result demonstrates that the calibrated model favors a more rapid growth in cholera cases than what is typically seen in the observed data, which explains why the model fails to predict the absence of cases between February, 2019 and October, 2022

Sus_df <- bake(
  file = "../model3/run_level_3/si_S_quants", {
    foreach(i=1:length(h3@times), .combine = bind_rows) %do% {
      get_h3_S_quants(bpf_out, i)
    }
  }
)

\begin{figure}[ht]
<<h3SusFig, echo=FALSE, message=FALSE, fig.height=3>>=
ggplot() +
  # Sims from initial condition
  geom_line(data = h3Spat_quants, aes(x = date, y = Q50), col = 'red') +
  geom_ribbon(
    data = h3Spat_quants,
    aes(x = date, ymin = Q025, ymax = Q975),
    fill = 'red', alpha = 0.5
  ) +
  # projections from initial condition, using filtered sims function
  geom_line(data = mod3_V0_initcond_res$SQuants, aes(x = date, y = Q50), col = 'red') +
  geom_ribbon(
    data = mod3_V0_initcond_res$SQuants,
    aes(x = date, ymin = Q025, ymax = Q975),
    fill = 'red', alpha = 0.5
  ) +
  # Sims from filtered distribution
  geom_line(
    data = Sus_df,
    aes(x = as.Date(lubridate::date_decimal(time)), y = Q50),
    col = 'blue'
  ) +
  geom_ribbon(
    data = Sus_df,
    aes(x = as.Date(lubridate::date_decimal(time)), ymin = Q025, ymax = Q975),
    fill = 'blue', alpha = 0.5
  ) +
  # Projections from filtered distribution
  geom_line(data = mod3_V0_res$SQuants, aes(x = date, y = Q50), col = 'blue') +
  geom_ribbon(
    data = mod3_V0_res$SQuants,
    aes(x = date, ymin = Q025, ymax = Q975),
    fill = 'blue', alpha = 0.5
  ) +
  facet_wrap(~dep, scales = 'free_y', nrow = 2) +
  theme_bw() +
  theme(axis.title.x = element_blank()) +
  ylab("Total Susceptible (%)")
@
\caption{\label{fig:h3Sus}Percentage of individuals that are in the susceptible compartment.
Simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.}
\end{figure}

>>>>

@@

@@

This measurement model differs from that used by \citet{lee20}, who fit the model in two stages: epidemic and endemic phases.
Although their text and supplement material do not explicitly describe the measurement model used, inspection of the code provided with their submission suggests a change in measurement model between the epidemic and endemic phases.
In the file \code{choleraEqsPublic.py}, \citet{lee20} create several functions, where each function represents a set of coupled differential equations that could potentially be used to model cholera incidence data.
Each function returns a vector (or \code{numpy} array) that represents the change in each state variable for a single time step, including the variable \code{dC}, which tracks the number of new infections and is used to obtain the reported case counts.
Following their comments and code, it appears that the function \code{choleraEqs10WithoutVaccinationNetwork} was used to describe the dynamics of the epidemic phase, and \code{choleraEqs11WithoutVaccinationNetwork} was used for the endemic stage (see Sec~\ref{sec:mod2rep} for more details).
Because their models were fit using least squares, the code in these functions suggest that the measurement model for the epidemic phase is
\begin{equation}
  \label{model2:measEpi}
  Y_{u, n} \mid \Delta N_{EI}(n) = z(n) \sim \text{N}\left(\reportRate z(n), \obsOverdispersion^2 \right),
\end{equation}
which is similar to our measurement model, the primary difference being that the measurement model is applied to raw case counts rather than log-transformed case counts.
The measurement model for the endemic phase modifies the epidemic model by counting both asymptomatically infected (A) and symptomatically infected (I) individuals in the case counts:
\begin{equation}
  \label{model2:measEnd}
  Y_{u, n} \mid \Delta N_{EI}(n) = z_1(n), \Delta N_{EA}(n) = z_2(n) \sim \text{N}\left(\reportRate \big(z_1(n) + z_2(n)\big), \obsOverdispersion^2 \right),
\end{equation}
where the notation for $\Delta N_{EA}(n)$ is similar to $\Delta N_{EI}(n)$, described above.

This measurement model is a minor change from that used by \cite{lee20}, which allowed for a change in the reporting rate on January 1st, 2018.
The fitted values of the reporting rate---before and after January 2018---were $0.97$ and $0.097$, respectively.
An instantaneous change from near perfect to almost non-existent reporting can be problematic, as it forces the model to explain the observed reduction in reported cases as a decrease in the reporting of cases, rather than a decrease in the prevalence of cholera.
This shift was justified by a "change of the case definition that occurred on January 1st, 2018";
this claim was not cited, and we could find no evidence that such a drastic change in the reporting rate would be warranted.
We therefore do not allow a change in reporting rate when fitting Model~3.

@@

As described in the main test, the latent states are initialized by estimating of the number of infected, asymptomatic, and recovered individuals at time $t_0$ by using observed case counts at time $t_1$.
The initial value for the bacteria compartment is then determined using the rainfall dynamics of Eq.~23.
For the departments of Grande'Anse, Nord-Ouest, and Sud-Est, estimating the latent states at time $t_0$ using the observed data at time $t_1$ is problematic (see Sec.~\ref{sec:mod3Cal}), so we instead estimate initial value parameters $I_{u0}(0)$ for the integers $u \in \{3, 7, 9\}$, which correspond to the aforementioned departments, and the remaining hidden states are estimated by enforcing model dynamics using this result.

@@

\section{Calibrating Model~3 to observed cases}\label{sec:mod3Cal}

# TODO: Move to run_level_3 folder...

In an attempt to determine why the model is underperforming in some departments, we plot the conditional log likelihoods of each observation given the previous observations for each department in Fig.~\ref{fig:condLL}.
This figure reveals that the fitted model has difficult time explaining certain features of the data.
For example, many departments (in particular Sud) have observations with lower conditional log-likelihoods near October 2016 than at other time points.
Further investigation of the model output reveals that the model is struggling to explain the sudden surge in cholera cases that occurred at this time, which coincides with the time that Hurricane Matthew struck Haiti.
While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \citep{ferreirai16}.
This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 that would increase the transmission between environmental cholera and humans for Sud and Grand'Anse.
The effect of the hurricane on cholera transmission is assumed to have an exponential decay, where the magnitude is determined by $\Whur{u}$ and the duration of the effect determined by $\hHur{u}$.

t(bpf_out@block.cond.loglik) %>%
  as.data.frame() %>%
  mutate(t = bpf_out@times) %>%
  pivot_longer(
    cols = -t,
    values_to = 'likelihood',
    names_prefix = "V",
    names_to = 'Unit'
  ) %>%
  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
  ggplot(aes(x = date, y = likelihood)) +
  geom_line() +
  geom_point() +
  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
             nrow = 2) +
  theme_bw() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle = -45, hjust = 0)) +
  ylab("Conditional log-likelihoods")
@
\caption{\label{fig:condLL}Conditional log-likelihoods of Model~3 prior to the inclusion of parameters for initial values and Hurricane Matthew.}
\end{figure}

Other points of concern include the observations at or near the start of the recorded cases, for example, in Nord-Ouest and Sud-Est.
These low conditional log-likelihoods are a result of having zero or near zero confirmed cases in the first week of observed data, followed by a rapid growth of cases.
The model cannot explain the rapid growth of cholera in the absence of starting cases because the initialization scheme forces departments with no recorded cases in the first week to have hidden states $I_{u0}(0) = A_{u0}(0) = B_{u}(0) = 0$.
In reality, however, it is possible to have some infected or asymptomatic individuals in a department when no cases are recorded due to under-reporting.
This may be particularly true at the onset of an infectious disease outbreak, as lower public awareness and inadequate testing infrastructure may result in many cases going undetected.
We therefore include and estimate $I_{0u}$ that determines the initial number of Infected individuals at the start of the pandemic for the departments Grande'Anse, Nord-Ouest, and Sud-Est;
each of these departments have both low starting case counts followed by rapid growth, and low initial values of the conditinoal log-likelihood.

\begin{figure}[ht]
<<finalCondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
t(bpf_out@block.cond.loglik) %>%
  as.data.frame() %>%
  mutate(t = bpf_out@times) %>%
  pivot_longer(
    cols = -t,
    values_to = 'likelihood',
    names_prefix = "V",
    names_to = 'Unit'
  ) %>%
  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
  ggplot(aes(x = date, y = likelihood)) +
  geom_line() +
  geom_point() +
  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
             nrow = 2) +
  theme_bw() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle = -45, hjust = 0)) +
  ylab("Conditional log-likelihoods")
@
\caption{\label{fig:finalCondLL}Conditional log-likelihoods of Model~3 after adding and estimating the parameters for initial values and Hurricane Matthew.}
\end{figure}

Recall that the filtering distribution for the calibrated version of Model~3 at time $t_k$ is defined as the distribution of the latent state at time $t_k$ given the data from times $t_{1}:t_{k}$, i.e. $f^{(3)}_{\bm{X}_k|\bm{Y}_{1:k}}(\bm{x}_{k} | \bm{y}^*_{1:k} ; \hat\theta)$.
In general, one may expect simulations from the filtering distribution of a model with a good statistical fit to result in hidden states that are highly consistent with the observed data because the filtering distribution is conditioned on the observed data.
Fig.~\ref{fig:h3Sus} shows the percentage of individuals that are in the susceptible compartment from various simulations of the model:
simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
This figure shows that simulations from initial conditions tends to result in a much more rapid depletion of susceptible individuals at the start of the epidemic than simulations from the filtering distribution, suggesting the calibrated model has a propensity to predict larger outbreaks than what is typically seen in the data.
This also appears true in later stages of the time period of observed data, where the data suggests that there is a replenishment of the susceptible compartment, but the Model simulations typically retain a smaller proportion of susceptible individuals.
This result demonstrates that the calibrated model favors a more rapid growth in cholera cases than what is typically seen in the observed data, which explains why the model fails to predict the absence of cases between February, 2019 and October, 2022.

Sus_df <- bake(
  file = "../model3/run_level_3/si_S_quants.rds", {
    foreach(i=1:length(h3@times), .combine = bind_rows) %do% {
      get_h3_S_quants(bpf_out, i)
    }
  }
)

\begin{figure}[ht]
<<h3SusFig, echo=FALSE, message=FALSE, fig.height=3>>=
ggplot() +
  # Sims from initial condition
  geom_line(data = h3Spat_quants, aes(x = date, y = Q50), col = 'red') +
  geom_ribbon(
    data = h3Spat_quants,
    aes(x = date, ymin = Q025, ymax = Q975),
    fill = 'red', alpha = 0.5
  ) +
  # projections from initial condition, using filtered sims function
  geom_line(data = mod3_V0_initcond_res$SQuants, aes(x = date, y = Q50), col = 'red') +
  geom_ribbon(
    data = mod3_V0_initcond_res$SQuants,
    aes(x = date, ymin = Q025, ymax = Q975),
    fill = 'red', alpha = 0.5
  ) +
  # Sims from filtered distribution
  geom_line(
    data = Sus_df,
    aes(x = as.Date(lubridate::date_decimal(time)), y = Q50),
    col = 'blue'
  ) +
  geom_ribbon(
    data = Sus_df,
    aes(x = as.Date(lubridate::date_decimal(time)), ymin = Q025, ymax = Q975),
    fill = 'blue', alpha = 0.5
  ) +
  # Projections from filtered distribution
  geom_line(data = mod3_V0_res$SQuants, aes(x = date, y = Q50), col = 'blue') +
  geom_ribbon(
    data = mod3_V0_res$SQuants,
    aes(x = date, ymin = Q025, ymax = Q975),
    fill = 'blue', alpha = 0.5
  ) +
  geom_vline(xintercept = max(h3Spat_quants$date), linetype = 'dashed') +
  facet_wrap(~dep, scales = 'free_y', nrow = 2) +
  theme_bw() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle = -45, hjust = 0)) +
  ylab("Total Susceptible (%)")
@
\caption{\label{fig:h3Sus}Percentage of individuals that are in the susceptible compartment.
Simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
The dashed line represents the end of the observed data.}
\end{figure}

@@

There are minor differences between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20sup}.
While the discrepancy appears minor, the deterministic nature of Model~2 implies that an exact replication of model trajectories should be possible.
In this case, these discrepancies may be attributed to implementing the model in two different programing languages.
Another potential explanation for the discrepancy is that the parameters that we used are only approximately the same as those used by \citet{lee20sup}.
For example, the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely had values that were far more precise.
Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.
The collective effect of these small differences in model parameters likely will result in small differences in model trajectories.

Some additional concerns about being able to accurately replicate the results of \cite{lee20} are valid.
Details about the measurement models and how latent states were initialized for the epidemic model were not provided by \citet{lee20sup} and therefore these details must be inferred by looking at the provided source code.
According to repository comments, the files \code{fit\allowbreak In\allowbreak Pieces\allowbreak 3params\allowbreak Clean\allowbreak May2019\allowbreak Public.py} and \code{fit\allowbreak In\allowbreak Pieces\allowbreak Mu\allowbreak With\allowbreak Frac\allowbreak Sus\allowbreak Fixed\allowbreak All\allowbreak Infections\allowbreak Public.py} were used to fit the epidemic and endemic phases of the model respectively, although it is apparent that these exact files were not used to obtain the reported results since the files contain some variable-naming errors that make it impossible to run the files without making modifications \footnote{One example of why the code cannot be run that the file loads functions from a non-extant file named \code{choleraEqs.py} in line 13 rather than \code{cholera\allowbreak Eqs\allowbreak Public.py}.}.
The inability to replicate the results by \citet{lee20} by running the provided source code makes checking whether or not a our numeric implementation faithfully represents their results very difficult.
Additionally, the script that was said to been used to obtain the results reported by \cite{lee20} appears to use a different measurement model than what was described in the supplemental material, again making it difficult to fully replicate the result of \cite{lee20} without being able to easily run the provided source code.
In this case, we chose to use measurement model described by Eq.~\ref{model2:measEpi} for both phases of the epidemic, as this seemed to visually match the results of \cite{lee20} most closely.

@@

lee3 <- lee3_spatPomp()
lee3_sims <- bake("../model3/run_level_3/si_lee3Sims.rds", {
  simulate(lee3, nsim = NSIM, seed = 321, format = 'data.frame')
})

@@

\section{Forecasting with parameter uncertainty}

Let $f_{Y_{1:N}}(y_{1:N} | \theta)$ denote the pdf of the model under consideration, were $\theta$ is a parameter vector that indexes the model.
Furthermore, denote the observed data as $y_{1:N}^*$.
Because the uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20},
parameter uncertainty should be considered when obtaining model forecasts when the goal is to influencing policy.
In a Bayesian modelling paradigm, the most natural way to account for parameter uncertainty in model forecasts is to suppose that $\theta$ comes from a distribution $f_{\Theta}$, and then to obtain $J$ forecasts from the model where each forecast is obtained using parameters drawn from the posterior distribution $\Theta_{1:J} \mid Y_{1:N} = y_{1:N}^* \sim f_{\Theta}\big(\theta | Y_{1:N} = y_{1:N}^*\big)$.

When frequentist methods are used, however, there does not exist a posterior distribution from which one could sample.
A common approach could be to obtain a weighted average of the simulations from various models \citep{hoeting99}, but this can be problematic when forecasts from each model are very different from each other \citep{grueber11}.
A similar approach that has been taken \citep{king15} is to obtain model forecasts using multiple sets of parameter values and then sample from the resulting forecasts using weights proportional to the corresponding likelihoods of the parameter values.
This approach could be considered a quasi-Empirical Bayes, as it is equivalent to using a discrete uniform prior where the set of values in the prior distribution is determined by a stochastic routine applied to the observed data, as discussed below.

For each $i \in 1:k$, let $\Theta_i$ be a random vector of model parameters.
Then, letting $\Theta$ denote the true model parameters, we endow the set $\{\Theta_1, \Theta_2, \ldots, \Theta_k\}$ with a discrete uniform distribution, such that $P\big(\Theta = \Theta_i\big) = \frac{1}{K}$ for all values $i \in \seq{1}{k}$.
Using this as a prior distribution, the posterior distribution of $\Theta | Y_{1:N} = y_{1:N}^*$ can be expressed as:
$P\big(\Theta = \Theta_k | Y_{1:N} = y_{1:N}^*\big) = \frac{f_{Y_{1:N}}(y_{1:N}^*| \Theta_k)}{\sum_{l = 1}^K f_{Y_{1:N}}(y_{1:N}^*| \theta_l)}$.
In a true Empirical Bayes analysis, the values $\Theta_1, \ldots, \Theta_k$ of the prior distribution would be chosen using the observed data, resulting in a posterior distribution that weighs the prior parameter vectors proportional to their corresponding likelihoods.
Instead, we choose $\Theta_i$ to be the output of a stochastic routine applied to the observed data by setting $\Theta_i$ to be the output of an iterated filtering algorithm.
In practice, because the likelihood maximization routines of iterated filtering methods are stochastic, it is common to run the iterated filtering method multiple times for each model in order to obtain a maximum likelihood estimate for model parameters.
This results in a natural set of parameters near the MLE that could be used in in the discrete prior distribution.
